
title: "Metabolomics Workshop: Regression & Classification with 1H NMR (RStudio)"
author: "A. Castillo"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
# Reproducibility
set.seed(123)

# Lightweight packages only (no tidyverse)
pkgs <- c(
  "caret",     # splitting, preprocessing
  "rpart",     # decision tree
  "rpart.plot",# tree plotting
  "class",     # kNN
  "pROC",      # (optional) ROC/AUC
  "ROCR"       # (optional) ROC curves
)
new <- pkgs[!(pkgs %in% installed.packages()[, "Package"])]
if (length(new)) install.packages(new, dependencies = TRUE)
for (p in pkgs) library(p, character.only = TRUE)
```

# 1. Dataset & Goal

**Goal**

- **Regression:** Use plasma 1H NMR spectra to predict numeric **BMI**.
- **Classification:** Use spectra to predict **COVID** (Positive) vs **No COVID**.

**Dataset**: `covid_19_noesy.tsv` (TSV).
- First columns (metadata): `id, category, Class, sex, age, BMI`.
- Remaining columns: spectral intensities (~8K points).
- Combine `Follow_up` + `Healthy_control` → **No COVID**; `Patient (+) SARS-CoV-2` → **COVID**.

We will:
1) Train/Test split (stratified).
2) PCA on spectra for dimensionality reduction.
3) Build/evaluate models (R²/MSE for regression; Confusion Matrix/Accuracy/F1 for classification).
4) (Optional) Try simple up/down-sampling on the **training set**.

---

# 2. Load & Prepare Data (Base R)

```{r load}
raw <- read.table(file="../2.univariate/covid_19_noesy.tsv", sep='\t', header=TRUE)


# Harmonize label: No COVID vs COVID
# Expect categories: "Follow_up", "Healthy_control", "Patient (+) SARS-CoV-2"
raw$covid_bin <- ifelse(raw$category %in% c("Follow_up", "Healthy_control"),
                        "No COVID",
                 ifelse(raw$category %in% c("Patient (+) SARS-CoV-2"),
                        "COVID", NA))
raw$covid_bin <- factor(raw$covid_bin, levels = c("No COVID", "COVID"))

cat("\nOriginal category counts:\n")
print(table(raw$category, useNA = "ifany"))
cat("\nBinary class counts (covid_bin):\n")
print(table(raw$covid_bin, useNA = "ifany"))
```

```{r select-spectra}
# Identify metadata columns present
meta_cols <- c("id", "category", "Class", "sex", "age", "BMI", "covid_bin")
meta_cols_present <- intersect(meta_cols, colnames(raw))

# Spectra columns = numeric columns not in metadata
num_idx  <- which(sapply(raw, is.numeric))
meta_idx <- match(meta_cols_present, colnames(raw))
meta_idx <- meta_idx[!is.na(meta_idx)]
spectra_idx <- setdiff(num_idx, meta_idx)

stopifnot(length(spectra_idx) > 50)  # expect many spectral points

spectra_mat <- as.matrix(raw[, spectra_idx, drop = FALSE])

# Keep a working frame with targets
Dat <- raw[, meta_cols_present, drop = FALSE]

# Quick summaries
if ("BMI" %in% colnames(Dat)) summary(Dat$BMI)
```

> **Note**: We intentionally use **only NMR spectra** (after PCA) as predictors for both tasks to keep the exercise focused on spectral learning. You can later add age/sex as extra predictors to compare.

---

# 3. Train/Test Split (Stratified on Class)

```{r split}
set.seed(123)
if (!("covid_bin" %in% colnames(Dat))) stop("covid_bin column missing after relabel.")

idx_cls <- createDataPartition(Dat$covid_bin, p = 0.7, list = FALSE)

train_spectra <- spectra_mat[idx_cls, , drop = FALSE]
train_meta    <- Dat[idx_cls, , drop = FALSE]

test_spectra  <- spectra_mat[-idx_cls, , drop = FALSE]
test_meta     <- Dat[-idx_cls, , drop = FALSE]

if ("BMI" %in% colnames(train_meta)) { summary(train_meta$BMI); summary(test_meta$BMI) }
```

---

3.5 Remove some regions from spectra

```{r roi}
# Convert spectral column names to numeric ppm values robustly
ppm_from_colnames <- function(nms) {
# Try direct numeric first (works if check.names = FALSE and names are like "-0.12", "4.70")
v <- suppressWarnings(as.numeric(nms))
if (sum(!is.na(v)) == length(nms)) return(v)
# Handle names like "X.0.095" (negative) and "X0.095" (positive)
out <- rep(NA_real_, length(nms))
is_neg <- grepl("^X\\.", nms)
is_pos <- grepl("^X(?!\\.)", nms, perl = TRUE)
out[is_neg] <- -suppressWarnings(as.numeric(sub("^X\\.", "", nms[is_neg])))
out[is_pos] <- suppressWarnings(as.numeric(sub("^X", "", nms[is_pos])))
# Fallback: try as.numeric raw
need <- is.na(out)
if (any(need)) out[need] <- suppressWarnings(as.numeric(nms[need]))
out
}


ppm <- ppm_from_colnames(colnames(spectra_mat))
# Sanity check: range of ppm
cat(sprintf("ppm range: [%.3f, %.3f]\n", min(ppm, na.rm = TRUE), max(ppm, na.rm = TRUE)))


# === Define ROI selection ===
# You can EITHER EXCLUDE ranges (e.g., water 4.5–5.1 ppm), or INCLUDE only a window.
# Use one mode at a time.
mode <- "exclude" # choices: "exclude" or "include"


# Example exclude ranges (modify as needed). Each element is c(min_ppm, max_ppm)
exclude_ranges <- list(c(-1, 2.8), c(4.5, 5.2)) # list(c(-1, 1.5), c(2, 3), c(4.5, 5.1), c(5.5, 6.5), c(8, 10)) # water region; add more ranges if needed
exclude_ranges <- list(c(4.5, 5.1)) # list(c(-1, 1.5), c(2, 3), c(4.5, 5.1), c(5.5, 6.5), c(8, 10)) # water region; add more ranges if needed


# Example include range (use only if mode == "include")
include_range <- c(-0.2, 0.2) # narrow ROI example around 0 ppm (modify as needed)


keep_idx <- rep(TRUE, length(ppm))
if (mode == "exclude" && length(exclude_ranges) > 0) {
for (rg in exclude_ranges) {
if (length(rg) == 2 && !any(is.na(rg))) {
lo <- min(rg); hi <- max(rg)
keep_idx <- keep_idx & !(ppm >= lo & ppm <= hi)
}
}
} else if (mode == "include" && length(include_range) == 2) {
lo <- min(include_range); hi <- max(include_range)
keep_idx <- (ppm >= lo & ppm <= hi)
}


# Apply mask and sort by decreasing ppm (conventional NMR axis)
ppm_kept <- ppm[keep_idx]
cols_kept <- which(keep_idx)
ord <- order(ppm_kept, decreasing = TRUE)


spectra_mat_roi <- spectra_mat[, cols_kept, drop = FALSE]
spectra_mat_roi <- spectra_mat_roi[, ord, drop = FALSE]
ppm_kept <- ppm_kept[ord]


# Quick overlay plot of all spectra in ROI
# (If many spectra, use transparency)
col_line <- grDevices::adjustcolor("black", alpha.f = 0.3)
matplot(ppm_kept, t(spectra_mat_roi), type = "l", lty = 1, col = col_line,
xlab = "ppm", ylab = "Intensity", main = "All spectra (ROI filtered)")


# Optional: visualize excluded ranges
if (mode == "exclude" && length(exclude_ranges) > 0) {
for (rg in exclude_ranges) {
lo <- min(rg); hi <- max(rg)
abline(v = c(lo, hi), col = "red", lty = 3)
}
}


# From here on, use the ROI-filtered spectra
spectra_mat <- spectra_mat_roi
```

---

# 4. PCA on Training Spectra, Apply to Test

```{r pca}
# PCA with centering & scaling
pca_fit <- prcomp(train_spectra, center = TRUE, scale. = TRUE)

# Variance explained to help choose PCs
var_explained <- pca_fit$sdev^2 / sum(pca_fit$sdev^2)
cumvar <- cumsum(var_explained)
head(data.frame(PC = seq_along(var_explained), Var = var_explained, CumVar = cumvar), 15)

# Choose number of PCs (e.g., 5). Adjust as needed (3–10)
nPC <- 5

# Project to PC space
train_PC <- pca_fit$x[, 1:nPC, drop = FALSE]
test_PC  <- predict(pca_fit, newdata = test_spectra)[, 1:nPC, drop = FALSE]

# For convenience, make data frames with targets
train_reg <- data.frame(BMI = train_meta$BMI, train_PC)
test_reg  <- data.frame(BMI = test_meta$BMI,  test_PC)

train_cls <- data.frame(covid_bin = train_meta$covid_bin, train_PC)
test_cls  <- data.frame(covid_bin = test_meta$covid_bin,  test_PC)
```

*Tip:* If the first ~5–10 PCs already capture a large fraction of variance (and show bio-separation), that’s usually enough for simple models.


# 6. CLASSIFICATION — COVID vs No COVID

We compare **Logistic Regression**, **Decision Tree**, and **k-NN** using the same PC features.

## 6.1 Logistic Regression
```{r logit}
# Logistic Regression (GLM with binomial)
logit_formula <- as.formula(paste("covid_bin ~", paste(colnames(train_PC), collapse = "+")))
logit_glm <- glm(logit_formula, data = data.frame(covid_bin = train_cls$covid_bin, train_PC), family = binomial)
summary(logit_glm)

# Predict probabilities on test
prob_logit <- predict(logit_glm, newdata = data.frame(covid_bin = test_cls$covid_bin, test_PC), type = "response")
pred_logit <- ifelse(prob_logit >= 0.5, "COVID", "No COVID")
pred_logit <- factor(pred_logit, levels = levels(test_cls$covid_bin))
```

## 6.1.2 Visualization of classification vs PCA

```{r logit-visuals, fig.width=7, fig.height=5}
## --- Helper plotting function ---
plot_pc_scatter <- function(PC, classes, title, main_suffix="") {
  # PC: matrix/data.frame with at least 2 columns (PC1, PC2)
  # classes: factor with class labels
  cols <- c("No COVID" = "steelblue", "COVID" = "tomato")
  
  plot(PC[,1], PC[,2], col = cols[as.character(classes)], pch = 19,
       xlab = "PC1", ylab = "PC2",
       main = paste(title, main_suffix))
  legend("topright", legend = levels(classes),
         col = cols, pch = 19, bty = "n")
}

## --- Training set: real vs predicted ---
# Real
plot_pc_scatter(train_PC, train_cls$covid_bin, "Training set", "— Real class")
# Predicted
prob_train <- predict(logit_glm, newdata = data.frame(train_PC), type = "response")
pred_train <- ifelse(prob_train >= 0.5, "COVID", "No COVID")
pred_train <- factor(pred_train, levels = levels(train_cls$covid_bin))
plot_pc_scatter(train_PC, pred_train, "Training set", "— Predicted class")

## --- Test set: real vs predicted ---
# Real
plot_pc_scatter(test_PC, test_cls$covid_bin, "Test set", "— Real class")
# Predicted (you already have pred_logit)
plot_pc_scatter(test_PC, pred_logit, "Test set", "— Predicted class")

```

## 6.2 Decision Tree
```{r tree}
set.seed(123)
# Use same predictor set (PCs) with rpart (classification)
tree_fit <- rpart(logit_formula, data = data.frame(covid_bin = train_cls$covid_bin, train_PC), method = "class",
                  control = rpart.control(cp = 0.01))
rpart.plot(tree_fit)

prob_tree <- predict(tree_fit, newdata = data.frame(covid_bin = test_cls$covid_bin, test_PC), type = "prob")[, "COVID"]
pred_tree <- ifelse(prob_tree >= 0.5, "COVID", "No COVID")
pred_tree <- factor(pred_tree, levels = levels(test_cls$covid_bin))
```
## Decision Tree with more splits
```{r}
set.seed(123)
tree_fit <- rpart(
  logit_formula,
  data = data.frame(covid_bin = train_cls$covid_bin, train_PC),
  method = "class",
  control = rpart.control(
    cp = 0.0001,      # very small cp → more splits
    minsplit = 5,     # allow splits with few samples
    minbucket = 2,    # allow small terminal nodes
    maxdepth = 5      # force deeper tree (adjust as you like)
  )
)

rpart.plot(tree_fit, type = 2, extra = 104, under = TRUE, faclen = 0)
```

## 6.3 k-Nearest Neighbors (kNN)
```{r knn}
# Prepare matrices for kNN
train_mat <- as.matrix(train_PC)
test_mat  <- as.matrix(test_PC)

# Standardize PCs (should already be scaled, but re-scale to be safe)
pre <- preProcess(train_mat, method = c("center", "scale"))
train_s <- predict(pre, train_mat)
test_s  <- predict(pre, test_mat)

k <- 5
pred_knn <- knn(train = train_s, test = test_s,
               cl = train_cls$covid_bin, k = k)
pred_knn <- factor(pred_knn, levels = levels(test_cls$covid_bin))
```

## 6.4 Evaluation: Confusion Matrix, Accuracy, Precision, Recall, F1
```{r metrics}
# Helper to compute metrics from predictions
compute_metrics <- function(truth, pred) {
  # Ensure both factors share the same levels
  lv <- levels(truth)
  truth_f <- factor(truth, levels = lv)
  pred_f  <- factor(pred,  levels = lv)
  cm <- table(Truth = truth_f, Pred = pred_f)
  acc <- sum(diag(cm)) / sum(cm)
  # Positive class = "COVID"
  tp <- cm["COVID", "COVID"]
  fp <- cm["No COVID", "COVID"]
  fn <- cm["COVID", "No COVID"]
  precision <- ifelse(tp + fp == 0, NA, tp / (tp + fp))
  recall    <- ifelse(tp + fn == 0, NA, tp / (tp + fn))
  f1 <- ifelse(is.na(precision) | is.na(recall) | (precision + recall == 0), NA,
               2 * precision * recall / (precision + recall))
  list(cm = cm, acc = acc, precision = precision, recall = recall, f1 = f1)
}

m_logit <- compute_metrics(test_cls$covid_bin, pred_logit)
m_tree  <- compute_metrics(test_cls$covid_bin, pred_tree)
m_knn   <- compute_metrics(test_cls$covid_bin, pred_knn)

# Show results
show_metrics <- function(name, m) {
  cat("\n===", name, "===\n")
  print(m$cm)
  cat(sprintf("Accuracy: %.3f\n", m$acc))
  cat(sprintf("Precision (COVID): %.3f\n", m$precision))
  cat(sprintf("Recall (COVID): %.3f\n", m$recall))
  cat(sprintf("F1 (COVID): %.3f\n", m$f1))
}

show_metrics("Logistic Regression", m_logit)
show_metrics("Decision Tree", m_tree)
show_metrics("kNN (k=5)", m_knn)

# Compact comparison table (base R)
comp <- data.frame(
  Model = c("Logistic", "Tree", "kNN (k=5)"),
  Accuracy = c(m_logit$acc, m_tree$acc, m_knn$acc),
  F1_COVID = c(m_logit$f1, m_tree$f1, m_knn$f1)
)
comp <- comp[order(-comp$F1_COVID), ]
print(comp, row.names = FALSE)
```

> **Why F1?** When classes are **imbalanced**, Accuracy can be misleading. **F1** balances Precision and Recall for the positive class (here, **COVID**).

---

# 7. (Optional) Addressing Class Imbalance via Resampling

We demonstrate **oversampling** on the **training set only** using ROSE. Re-train Logistic Regression on the resampled training data.

```{r resample, eval=FALSE}
# Only run if you want to try oversampling
if (!require(ROSE)) install.packages("ROSE"); library(ROSE)

train_aug <- data.frame(covid_bin = train_cls$covid_bin, train_PC)

# ROSE synthetic oversampling towards balanced classes
set.seed(123)
rose_out <- ROSE(covid_bin ~ ., data = train_aug, seed = 123)$data

print(table(train_cls$covid_bin))
print(table(rose_out$covid_bin))

# Refit logistic on oversampled data
logit_glm_ros <- glm(covid_bin ~ ., data = rose_out, family = binomial)
prob_logit_ros <- predict(logit_glm_ros, newdata = data.frame(covid_bin = test_cls$covid_bin, test_PC), type = "response")
pred_logit_ros <- ifelse(prob_logit_ros >= 0.5, "COVID", "No COVID")
pred_logit_ros <- factor(pred_logit_ros, levels = levels(test_cls$covid_bin))

m_logit_ros <- compute_metrics(test_cls$covid_bin, pred_logit_ros)
show_metrics("Logistic (ROSE Oversampled Train)", m_logit_ros)
```

> **Guideline:** Always apply resampling **only on the training set**, never on the test set.

---

# 8. Quick Visuals for Interpretation (Base Graphics)

```{r visuals, fig.width=7, fig.height=5}
# PCA scores plot (first two PCs) colored by class
cols <- as.integer(train_cls$covid_bin)
plot(train_PC[,1], train_PC[,2], col = cols,
     xlab = "PC1", ylab = "PC2", main = "Training PCA Scores by Class")
legend("topright", legend = levels(train_cls$covid_bin),
       col = 1:length(levels(train_cls$covid_bin)), pch = 1)

# Logistic coefficients (by PC)
co <- coef(summary(logit_glm))
# Drop intercept if present
if ("(Intercept)" %in% rownames(co)) co <- co[setdiff(rownames(co), "(Intercept)"), , drop = FALSE]
barplot(co[, "Estimate"], names.arg = rownames(co), las = 2,
        main = "Logistic Regression Coefficients (PCs)", ylab = "Estimate")
```

---

# 9. What to Try Next (Exercises)

1. Vary `nPC` between 3–10. How do R²/MSE (regression) and F1 (classification) change?
2. Add clinical covariates (e.g., **age**, **sex**) to the PC predictors. Does performance improve? Any overfitting?
3. Tune decision tree complexity (`cp`) and kNN `k` (try 3, 7, 9...).
4. Try **undersampling** the majority class (e.g., with ROSE `ovun.sample(method = "under")`).
5. Create ROC curves and compute AUC (packages `pROC` or `ROCR`).

---

# 10. Appendix: Session Info

```{r session}
sessionInfo()
```

