
title: "Metabolomics Workshop: Regression & Classification with 1H NMR (RStudio)"
author: "A. Castillo"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
# Reproducibility
set.seed(123)

# Lightweight packages only (no tidyverse)
pkgs <- c(
  "caret",     # splitting, preprocessing
  "rpart",     # decision tree
  "rpart.plot",# tree plotting
  "class",     # kNN
  "pROC",      # (optional) ROC/AUC
  "ROCR"       # (optional) ROC curves
)
new <- pkgs[!(pkgs %in% installed.packages()[, "Package"])]
if (length(new)) install.packages(new, dependencies = TRUE)
for (p in pkgs) library(p, character.only = TRUE)
```

# 1. Dataset & Goal

**Goal**

- **Regression:** Use plasma 1H NMR spectra to predict numeric **BMI**.
- **Classification:** Use spectra to predict **COVID** (Positive) vs **No COVID**.

**Dataset**: `covid_19_noesy.tsv` (TSV).
- First columns (metadata): `id, category, Class, sex, age, BMI`.
- Remaining columns: spectral intensities (~8K points).
- Combine `Follow_up` + `Healthy_control` → **No COVID**; `Patient (+) SARS-CoV-2` → **COVID**.

We will:
1) Train/Test split (stratified).
2) PCA on spectra for dimensionality reduction.
3) Build/evaluate models (R²/MSE for regression; Confusion Matrix/Accuracy/F1 for classification).
4) (Optional) Try simple up/down-sampling on the **training set**.

---

# 2. Load & Prepare Data (Base R)

```{r load}
#raw <- read.table(file="../2.univariate/covid_19_cpmg.tsv", sep='\t', header=TRUE)
raw <- read.table(file="../2.univariate/covid_19_noesy.tsv", sep='\t', header=TRUE)


# Harmonize label: No COVID vs COVID
# Expect categories: "Follow_up", "Healthy_control", "Patient (+) SARS-CoV-2"
raw$covid_bin <- ifelse(raw$category %in% c("Follow_up", "Healthy_control"),
                        "No COVID",
                 ifelse(raw$category %in% c("Patient (+) SARS-CoV-2"),
                        "COVID", NA))
raw$covid_bin <- factor(raw$covid_bin, levels = c("No COVID", "COVID"))

cat("\nOriginal category counts:\n")
print(table(raw$category, useNA = "ifany"))
cat("\nBinary class counts (covid_bin):\n")
print(table(raw$covid_bin, useNA = "ifany"))
```

```{r}
# Create a barchar with the counts of each category
barplot(table(raw$category), col = c("lightblue", "salmon"),
        main = "Counts by category", ylab = "Count", xlab = "Category")  
  
```


```{r select-spectra}
# Identify metadata columns present
meta_cols <- c("id", "category", "Class", "sex", "age", "BMI", "covid_bin")
meta_cols_present <- intersect(meta_cols, colnames(raw))

# Remove rows where BMI <= 0
raw <- raw[raw$BMI > 0,]

# Spectra columns = numeric columns not in metadata
num_idx  <- which(sapply(raw, is.numeric))
meta_idx <- match(meta_cols_present, colnames(raw))
meta_idx <- meta_idx[!is.na(meta_idx)]
spectra_idx <- setdiff(num_idx, meta_idx)

stopifnot(length(spectra_idx) > 50)  # expect many spectral points

# Remove rows where BMI <= 0
spectra_mat <- as.matrix(raw[, spectra_idx, drop = FALSE])

# Keep a working frame with targets
Dat <- raw[, meta_cols_present, drop = FALSE]

# Quick summaries
if ("BMI" %in% colnames(Dat)) summary(Dat$BMI)
```

> **Note**: We intentionally use **only NMR spectra** (after PCA) as predictors for both tasks to keep the exercise focused on spectral learning. You can later add age/sex as extra predictors to compare.

---

# 3. Train/Test Split (Stratified on Class)

```{r split}
set.seed(123)
if (!("covid_bin" %in% colnames(Dat))) stop("covid_bin column missing after relabel.")

idx_cls <- createDataPartition(Dat$covid_bin, p = 0.7, list = FALSE)

train_spectra <- spectra_mat[idx_cls, , drop = FALSE]
train_meta    <- Dat[idx_cls, , drop = FALSE]

test_spectra  <- spectra_mat[-idx_cls, , drop = FALSE]
test_meta     <- Dat[-idx_cls, , drop = FALSE]

if ("BMI" %in% colnames(train_meta)) { summary(train_meta$BMI); summary(test_meta$BMI) }
```

---

# 4. PCA on Training Spectra, Apply to Test

```{r pca}
# PCA with centering & scaling
pca_fit <- prcomp(train_spectra, center = TRUE, scale. = TRUE)

# Variance explained to help choose PCs
var_explained <- pca_fit$sdev^2 / sum(pca_fit$sdev^2)
cumvar <- cumsum(var_explained)
head(data.frame(PC = seq_along(var_explained), Var = var_explained, CumVar = cumvar), 15)

# Choose number of PCs (e.g., 5). Adjust as needed (3–10)
nPC <- 5

# Project to PC space
train_PC <- pca_fit$x[, 1:nPC, drop = FALSE]
test_PC  <- predict(pca_fit, newdata = test_spectra)[, 1:nPC, drop = FALSE]

# For convenience, make data frames with targets
train_reg <- data.frame(BMI = train_meta$BMI, train_PC)
test_reg  <- data.frame(BMI = test_meta$BMI,  test_PC)

train_cls <- data.frame(covid_bin = train_meta$covid_bin, train_PC)
test_cls  <- data.frame(covid_bin = test_meta$covid_bin,  test_PC)
```

*Tip:* If the first ~5–10 PCs already capture a large fraction of variance (and show bio-separation), that’s usually enough for simple models.

---

# 5. REGRESSION — Predict BMI from PCs

```{r regression-fit}
# Multiple Linear Regression (BMI ~ PCs)
reg_formula <- as.formula(paste("BMI ~", paste(colnames(train_PC), collapse = "+")))
reg_lm <- lm(reg_formula, data = train_reg)
summary(reg_lm)
```

```{r regression-eval 0}
# Predict on test
pred_bmi <- predict(reg_lm, newdata = train_reg)

gt <- train_reg$BMI
mse <- mean((gt - pred_bmi)^2)
ss_res <- sum((gt - pred_bmi)^2)
ss_tot <- sum((gt - mean(gt, na.rm = TRUE))^2)
r2 <- 1 - ss_res/ss_tot

cat(sprintf("Train MSE: %.3f\n", mse))
cat(sprintf("Train R^2: %.3f\n", r2))

# Plot predicted vs. actual
plot(gt, pred_bmi,
     xlab = "Actual BMI", ylab = "Predicted BMI",
     main = sprintf("BMI Regression: Train Set (nPC=%d)\nR^2=%.3f, MSE=%.3f", nPC, r2, mse))
abline(0, 1, lty = 2)

```


```{r regression-eval}
# Predict on test
pred_bmi <- predict(reg_lm, newdata = test_reg)

gt <- test_reg$BMI
mse <- mean((gt - pred_bmi)^2)
ss_res <- sum((gt - pred_bmi)^2)
ss_tot <- sum((gt - mean(gt, na.rm = TRUE))^2)
r2 <- 1 - ss_res/ss_tot

cat(sprintf("Test MSE: %.3f\n", mse))
cat(sprintf("Test R^2: %.3f\n", r2))

# Plot predicted vs. actual
plot(gt, pred_bmi,
     xlab = "Actual BMI", ylab = "Predicted BMI",
     main = sprintf("BMI Regression: Test Set (nPC=%d)\nR^2=%.3f, MSE=%.3f", nPC, r2, mse))
abline(0, 1, lty = 2)
```

> **Interpretation:** Higher R² and lower MSE indicate better predictive performance. Check `summary(reg_lm)` to see which PCs contribute most.

---

That clearly does not work. So, let's clean the data first


```{r roi}
# Convert spectral column names to numeric ppm values robustly
ppm_from_colnames <- function(nms) {
# Try direct numeric first (works if check.names = FALSE and names are like "-0.12", "4.70")
v <- suppressWarnings(as.numeric(nms))
if (sum(!is.na(v)) == length(nms)) return(v)
# Handle names like "X.0.095" (negative) and "X0.095" (positive)
out <- rep(NA_real_, length(nms))
is_neg <- grepl("^X\\.", nms)
is_pos <- grepl("^X(?!\\.)", nms, perl = TRUE)
out[is_neg] <- -suppressWarnings(as.numeric(sub("^X\\.", "", nms[is_neg])))
out[is_pos] <- suppressWarnings(as.numeric(sub("^X", "", nms[is_pos])))
# Fallback: try as.numeric raw
need <- is.na(out)
if (any(need)) out[need] <- suppressWarnings(as.numeric(nms[need]))
out
}


ppm <- ppm_from_colnames(colnames(spectra_mat))
# Sanity check: range of ppm
cat(sprintf("ppm range: [%.3f, %.3f]\n", min(ppm, na.rm = TRUE), max(ppm, na.rm = TRUE)))


# === Define ROI selection ===
# You can EITHER EXCLUDE ranges (e.g., water 4.5–5.1 ppm), or INCLUDE only a window.
# Use one mode at a time.
mode <- "exclude" # choices: "exclude" or "include"


# Example exclude ranges (modify as needed). Each element is c(min_ppm, max_ppm)
exclude_ranges <- list(c(-1, 2.8), c(4.5, 5.2)) # list(c(-1, 1.5), c(2, 3), c(4.5, 5.1), c(5.5, 6.5), c(8, 10)) # water region; add more ranges if needed
exclude_ranges <- list(c(4.5, 5.1)) # list(c(-1, 1.5), c(2, 3), c(4.5, 5.1), c(5.5, 6.5), c(8, 10)) # water region; add more ranges if needed


# Example include range (use only if mode == "include")
include_range <- c(-0.2, 0.2) # narrow ROI example around 0 ppm (modify as needed)


keep_idx <- rep(TRUE, length(ppm))
if (mode == "exclude" && length(exclude_ranges) > 0) {
for (rg in exclude_ranges) {
if (length(rg) == 2 && !any(is.na(rg))) {
lo <- min(rg); hi <- max(rg)
keep_idx <- keep_idx & !(ppm >= lo & ppm <= hi)
}
}
} else if (mode == "include" && length(include_range) == 2) {
lo <- min(include_range); hi <- max(include_range)
keep_idx <- (ppm >= lo & ppm <= hi)
}


# Apply mask and sort by decreasing ppm (conventional NMR axis)
ppm_kept <- ppm[keep_idx]
cols_kept <- which(keep_idx)
ord <- order(ppm_kept, decreasing = TRUE)


spectra_mat_roi <- spectra_mat[, cols_kept, drop = FALSE]
spectra_mat_roi <- spectra_mat_roi[, ord, drop = FALSE]
ppm_kept <- ppm_kept[ord]


# Quick overlay plot of all spectra in ROI
# (If many spectra, use transparency)
col_line <- grDevices::adjustcolor("black", alpha.f = 0.3)
matplot(ppm_kept, t(spectra_mat_roi), type = "l", lty = 1, col = col_line,
xlab = "ppm", ylab = "Intensity", main = "All spectra (ROI filtered)")


# Optional: visualize excluded ranges
if (mode == "exclude" && length(exclude_ranges) > 0) {
for (rg in exclude_ranges) {
lo <- min(rg); hi <- max(rg)
abline(v = c(lo, hi), col = "red", lty = 3)
}
}


# From here on, use the ROI-filtered spectra
spectra_mat <- spectra_mat_roi
```
---

# 3. Train/Test Split (Stratified on Class)




```{r split}
set.seed(123)
if (!("covid_bin" %in% colnames(Dat))) stop("covid_bin column missing after relabel.")

idx_cls <- createDataPartition(Dat$covid_bin, p = 0.7, list = FALSE)

train_spectra <- spectra_mat[idx_cls, , drop = FALSE]
train_meta <- Dat[idx_cls, , drop = FALSE]

test_spectra <- spectra_mat[-idx_cls, , drop = FALSE]
test_meta <- Dat[-idx_cls, , drop = FALSE]

if ("BMI" %in% colnames(train_meta)) { summary(train_meta$BMI); summary(test_meta$BMI) }
```
---


# 4. PCA on Training Spectra, Apply to Test


```{r pca}
# PCA with centering & scaling
#pca_fit <- prcomp(train_spectra, center = TRUE, scale. = TRUE)
#pca_fit <- prcomp(train_spectra, center = FALSE, scale. = FALSE)



# Variance explained to help choose PCs
var_explained <- pca_fit$sdev^2 / sum(pca_fit$sdev^2)
cumvar <- cumsum(var_explained)
head(data.frame(PC = seq_along(var_explained), Var = var_explained, CumVar = cumvar), 15)


# Choose number of PCs (e.g., 5). Adjust as needed (3–10)
nPC <- 38


# Project to PC space
train_PC <- pca_fit$x[, 1:nPC, drop = FALSE]
test_PC <- predict(pca_fit, newdata = test_spectra)[, 1:nPC, drop = FALSE]


# For convenience, make data frames with targets
train_reg <- data.frame(BMI = train_meta$BMI, train_PC)
test_reg <- data.frame(BMI = test_meta$BMI, test_PC)


train_cls <- data.frame(covid_bin = train_meta$covid_bin, train_PC)
test_cls <- data.frame(covid_bin = test_meta$covid_bin, test_PC)

```


*Tip:* If the first ~5–10 PCs already capture a large fraction of variance (and show bio-separation), that’s usually enough for simple models.

---

```{r plot_pca}
## PC1 vs PC2 colored by BMI (training set) — base R, no tidyverse

# If you already computed PCA -> ensure train_PC has at least 2 PCs.
# Otherwise compute PCA on train_spectra:
if (!exists("train_PC")) {
  stopifnot(exists("train_spectra"))
  pca_fit  <- prcomp(train_spectra, center = FALSE, scale. = FALSE)
  train_PC <- pca_fit$x  # rows = samples, cols = PCs
}

# Grab BMI (numeric) for training samples
stopifnot(exists("train_meta"))
bmi <- as.numeric(train_meta$BMI)

# Drop NAs if any
ok <- is.finite(bmi) & is.finite(train_PC[,1]) & is.finite(train_PC[,2])
PC1 <- train_PC[ok, 1]
PC2 <- train_PC[ok, 2]
bmi_ok <- bmi[ok]

# Map BMI to a color gradient
pal <- colorRampPalette(c("navy", "cyan", "yellow", "red"))
nbins <- 100
cols  <- pal(nbins)
zcut  <- as.integer(cut(bmi_ok, breaks = nbins, include.lowest = TRUE))

# Plot
plot(PC1, PC2,
     col = cols[zcut], pch = 19,
     xlab = "PC1", ylab = "PC2",
     main = "Training: PC1 vs PC2 colored by BMI")

# Simple continuous color bar (right side)
usr <- par("usr")
x0 <- usr[2] + 0.04 * diff(usr[1:2])
x1 <- x0 + 0.03 * diff(usr[1:2])
y0 <- usr[3]; y1 <- usr[4]
rect(xleft = x0, ybottom = seq(y0, y1, length.out = nbins)[-nbins],
     xright = x1, ytop = seq(y0, y1, length.out = nbins)[-1],
     col = cols, border = NA)
# Add ticks/labels
axis_y_vals <- pretty(range(bmi_ok), n = 5)
axis_y_pos  <- y0 + (axis_y_vals - min(bmi_ok)) / diff(range(bmi_ok)) * (y1 - y0)
segments(x1, axis_y_pos, x1 + 0.01 * diff(usr[1:2]), axis_y_pos)
text(x1 + 0.015 * diff(usr[1:2]), axis_y_pos, labels = axis_y_vals, adj = 0)
text(x0, y1 + 0.04 * diff(usr[3:4]), labels = "BMI", adj = c(0, 0))

# Optional: add a legend for number of points
legend("topright", legend = sprintf("n = %d", length(bmi_ok)), bty = "n")
```
```{r}
# Plot the explained variance for the previous PCA
plot(cumvar, type = 'b', pch = 19, xlab = "Number of PCs", ylab = "Cumulative Variance Explained",
     main = "PCA Variance Explained")
abline(v = nPC, col = "red", lty = 2) 
text(nPC + 1, 0.5, labels = sprintf("nPC = %d", nPC), col = "red", adj = c(0, 0))   

```

---


# 5. REGRESSION — Predict BMI from PCs


```{r regression-fit}
# Multiple Linear Regression (BMI ~ PCs)
reg_formula <- as.formula(paste("BMI ~", paste(colnames(train_PC), collapse = "+")))
reg_lm <- lm(reg_formula, data = train_reg)
summary(reg_lm)

```

```{r regression-eval}
# Predict on test
pred_bmi <- predict(reg_lm, newdata = train_reg)


gt <- train_reg$BMI
mse <- mean((gt - pred_bmi)^2)
ss_res <- sum((gt - pred_bmi)^2)
ss_tot <- sum((gt - mean(gt, na.rm = TRUE))^2)
r2 <- 1 - ss_res/ss_tot


cat(sprintf("Train MSE: %.3f", mse))
cat(sprintf("Train R^2: %.3f", r2))


# Plot predicted vs. actual
plot(gt, pred_bmi,
xlab = "Actual BMI", ylab = "Predicted BMI",
main = sprintf("BMI Regression: Train Set (nPC=%d)
R^2=%.3f, MSE=%.3f", nPC, r2, mse))
abline(0, 1, lty = 2)

```

```{r regression-eval}
# Predict on test
pred_bmi <- predict(reg_lm, newdata = test_reg)


gt <- test_reg$BMI
mse <- mean((gt - pred_bmi)^2)
ss_res <- sum((gt - pred_bmi)^2)
ss_tot <- sum((gt - mean(gt, na.rm = TRUE))^2)
r2 <- 1 - ss_res/ss_tot


cat(sprintf("Test MSE: %.3f
", mse))
cat(sprintf("Test R^2: %.3f
", r2))


# Plot predicted vs. actual
plot(gt, pred_bmi,
xlab = "Actual BMI", ylab = "Predicted BMI",
main = sprintf("BMI Regression: Test Set (nPC=%d)
R^2=%.3f, MSE=%.3f", nPC, r2, mse))
abline(0, 1, lty = 2)

```



> **Interpretation:** Higher R² and lower MSE indicate better predictive performance. Check `summary(reg_lm)` to see which PCs contribute most.


