---
title: "Getting Started with R Notebooks — A Gentle Bioinformatics Intro"
author: "Your Name"
output: html_notebook
---

# Welcome

Welcome to your first R Notebook!  

In this session, we will:

1. Learn how R Notebooks mix **code** and **text**.
2. Create some **fake bioinformatics-style data**.
3. Explore and inspect a dataframe.
4. Make a simple scatter plot.

💡 *Tip: Press `Shift + Enter` to run the current cell.*
  
---

## 1. Hello World

```{r}
# This is a code cell
# Lines starting with '#' are comments and won't run as code
# Let's print a friendly message
print("Hello, Bioinformatics world!")
```

---

## 2. Creating Fake Data

```{r}
# Let's create some fake data that could look like metabolomics results

# Set a seed for reproducibility
set.seed(123)

# Create a dataframe with sample IDs, concentration values, and fold change
bio_data <- data.frame(
  SampleID = paste0("Sample_", 1:10),
  Concentration = round(rnorm(10, mean = 50, sd = 5), 2), # Random normal values
  FoldChange = round(rnorm(10, mean = 1.5, sd = 0.3), 2)  # Random normal values
)

# View the first rows
head(bio_data)
```

---

## 3. Inspecting the Dataframe

```{r}
# Check the structure of the dataframe
str(bio_data)

# Get summary statistics
summary(bio_data)

# See the names of the columns
colnames(bio_data)
```

---

## 4. Notes on Data Inspection

- **`str()`**: Shows the structure (column names, types, first few values).
- **`summary()`**: Gives descriptive statistics.
- **`colnames()`**: Lists the column names.

---

## 5. Scatter Plot

```{r}
# Let's make a scatter plot of Concentration vs FoldChange

plot(
  bio_data$Concentration,
  bio_data$FoldChange,
  main = "Concentration vs Fold Change",
  xlab = "Concentration (arbitrary units)",
  ylab = "Fold Change",
  pch = 19, # solid circles
  col = "blue"
)

# Optional: Add sample labels
text(
  bio_data$Concentration,
  bio_data$FoldChange,
  labels = bio_data$SampleID,
  pos = 4, cex = 0.8
)
```

---

## 6. Wrap-Up

✅ How to run code cells and markdown cells  
✅ How to create a dataframe in R  
✅ How to inspect and summarize data  
✅ How to make a simple scatter plot  

Next steps:  
In the next lesson, we will explore how to use R packages like **BioConductor** to analyze real biological datasets.
