---
title: "Correlation Analysis of Simulated BioResults"
author: "Your Name"
output: html_notebook
---

# Introduction

In this notebook, we will:

1. Load a CSV file containing simulated bioresults.
2. Inspect the data.
3. Create a scatter plot to visualize the correlation between two variables.

---

## 1. Load the CSV File

```{r}
# Load the CSV file
bio_df <- read.csv("correlated_bioresults.csv")

# Inspect the first few rows
head(bio_df)
```

---

## 2. Scatter Plot with Correlation

```{r}
# Scatter plot
plot(
  bio_df$GeneExpression,
  bio_df$ProteinLevel,
  main = "Correlation between Gene Expression and Protein Level",
  xlab = "Gene Expression (a.u.)",
  ylab = "Protein Level (a.u.)",
  pch = 19,
  col = "darkgreen"
)

# Add a linear regression line
abline(lm(ProteinLevel ~ GeneExpression, data = bio_df), col = "red", lwd = 2)

# Calculate correlation coefficient
cor_value <- cor(bio_df$GeneExpression, bio_df$ProteinLevel)

# Show correlation value in legend
legend(
  "topleft",
  legend = paste("Correlation:", round(cor_value, 2)),
  bty = "n"
)
```

---

## 3. Conclusion

The plot shows the relationship between **Gene Expression** and **Protein Level** for our simulated dataset.
The calculated correlation coefficient provides a quantitative measure of how strongly these two variables are related.
